SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-31 2017-Xudong-SMT/ActaInformatica.tex
some typos corrected

pNets can encode naturally the notion of input actions as found,
\eg in value-passing CCS 
\cite{Milner89} or of usual point-to-point message passing calculi,
but it also allows 
for more general mechanisms, like gate negotiation in Lotos~\cite{LotosISO89}, or broadcast
communications.

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$
          Add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{OA_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$; \Comment Fixpoint reached
\State \Return $\mathit{OA_n}$;

@@

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
\State $\mathit{L2}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L2)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{OA_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$;
\State \Return $\mathit{OA_n}$;
%\EndIf
\end{algorithmic}  
\end{algorithm}

@@

 Let $pn=$ be a pNet. The set of states 
 managed by the two algorithms both are subsets of the cartesian
 product of the pLTS states, as defined in page \pageref{def-states}. In a
 similar way, the set of state variables attached to each state are
 also equal. Comparing  
 transitions requires a little more care, because within the
 MakeTransitions function, fresh variables are created everytime a
 synchronisation vector is used (this occurs within the $MkPred$ and
 $matching$ functions). However, given a global state, the call of
 $MakeTransitions$ will always returns exactly the same set of OTs, up to
renaming of these fresh variables. When this is not ambiguous, we will
speak of equality (of transitions, and of open automata) when we
should say ``up to renaming of fresh variables''.

  Let us first define some notations.
 \begin{itemize}
    \item $L0$: the original set of transitions by Algorithm 1
      (i.e. before line 11), 
    \item $L1_{sat}$:  the original set of transitions by Algorithm 1, after
      satisfiability check (i.e. after line 11),
    \item $L1_{reach}$: the final set of OTs of Algorithm 1, as computed
      by line 12 (as $OA_0$). Without assignments the following
      refinement step will do nothing, the fixpoint is already reached.
    \item $L2_{sat}$: the original set of transitions by Algorithm 3
      (i.e. before line 13), 
    \item $L2_{reach}$: the final set of OTs of Algorithm 2, as computed
          by line 13. Here also $OA_n = OA_0$.
\end{itemize}

\paragraph{States:} for each of these, define
 $States: \overline{OT}\rightarrow\overline{State}$ the function
computing the set of target states of a set of OTs. \\
\noindent
\emph{Transitions:} between two sets of open transitions computed
by the algorithms, define $Rename:  L1_{reach} \rightarrow L2_{reach}$
which is the function renaming fresh variables.

%% When the pNets contains no assignment, there is no need for the last
%% filtering step, so $L1_{sat}=L1_{reach}$ and $L2_{sat}=L2_{reach}$.
We want to prove that $L1_{reach}=L2_{reach}$, that will imply
the equality of the 2 open automata.

\smallskip
Let $S=\ostate{\overline{s}}=\ostate{\{s_{i}\}^{i\in Leaves(pn)}}$
range over the global states $States(L0)$, 
The only difference between the main loops of the algorithms is that 
algorithm~\ref{alg3} only retains the satisfiable OTs (line 6-7), so 
algorithm~\ref{alg3} computes less transitions than
algorithm~\ref{alg1}. Formaly $L2_{reach}\subseteq  L2_{sat} \subseteq L0$, and as a consequence $States(L2_{reach})\subseteq  States(L0)$, so we have:\\
\centerline{$\forall \ostate{\overline{u}}\in States(L2_{reach}), \exists\ostate{\overline{t}}\in States(L0), \ostate{\overline{t}}=\ostate{\overline{u}}$}

For a global state $S$, and a set of OTs $L$, define its distance
$|S|_L$ as the length $n$ of (one of the) shortest chain
$S_0\xrightarrow{l_1}S_1\xrightarrow{l_2}...S_{n-1}\xrightarrow{l_{n}}S$
  from the initial state $S_0$, with $\forall k\in [1..n], l_k \in L$.

\medskip
We will prove by induction on the distance from the original state,
that $\forall S \in L0. \forall n \in \mathbb{N}. S \in States(L2_{reach}) \land |S|_{L2_{reach}}=n \Rightarrow S
\in States(L1_{reach})$, and vice-versa.

@@

  Suppose our property is true for distance $n$, and
  suppose $|S|_{L2_{reach}}=n+1$, then there exists a chain
  $S_0\xrightarrow{l_1}S_1\xrightarrow{l_2}...S_{n}\xrightarrow{l_{n+1}}S$
  with all $l_k \in L2_{reach}$. In particular, as
  $L2_{reach}\subseteq L2_{sat}$, $l_{n+1} \in L2_{sat}$, meaning 
  its predicate is satisfiable.

  By induction hypothesis, we get $S_{n} \in States(L1_{reach})$, so during the
  call of $makeReachableSubAutomaton$ in algorithm~\ref{alg1}, line 12, the OT
  $S_n \xrightarrow{l_{n}}S$ will be check and found SAT, so
  $S \in States(L1_{reach})$.

  Conversely if $|S|_{L1_{reach}}=n+1$, we get a similar chain, in
  $L1_{reach}$, with $S_n \xrightarrow{l_{n}}S, S_n \in
  States(L1_{reach})$, and $l_n$ satisfiable. By induction, we have
  $S_n \in States(L2_{reach})=States(L2_{sat})$, so $l_n$ will have
  been checked ``on-the-fly'', in line 6 of algorithm~\ref{alg3},
  the OT $S_n \xrightarrow{l_{n}}S$ added to $L2_{sat}$, and we
  conclude $S \in States(L2_{reach})$.

Remember $L0$ is finite, so we conclude $L1_{reach}=L2_{reach}$, and $States(L1_{reach})=States(L2_{reach})$.

@@

\paragraph{General case with assignments.}
In the presence of assignments, we need to collect all of them
  before using the assignment information to refine the SAT
  checking. Still, the smart strategy will explore a reduced state
  space, not necessarily the minimal one. In the
  following theorem, both algorithms are considered in their complete
  version, meaning the reachability (with assignments) step is
  iterated up to fixpoint.

\begin{theorem}[Correctness with assignments]
  \label{Th:CorrectnessWithAssignments}
In the general case, if a pNet (verifying the finiteness conditions of theorem 1), contains some assignments to state variables, the smart algorithm computes 
the same open automaton as that of Algorithm 1 up to a renaming of variables in the transition labels.
\end{theorem}

%\TODO{Formalize the proof}
\begin{proof}
  Define the OT sets $L0, L1_{sat}, L2_{sat}, L2_{reach}$ as previously, and:

  \begin{itemize}
    \item $L1_{reach*}$ = final set of OTs of Algorithm 1, as computed
        by an iterative application of makeReachableSubAutomaton (line
        14) up to fixpoint.
      \item $L2_{reach*}$ = final set of OTs of Algorithm 2, as computed
        by an iterative application of makeReachableSubAutomaton (line
        16) up to fixpoint.
  \end{itemize}

?? In Theorem 2, we have already proved that the result automata $OA_1$ and $OA_2$ of two algorithms are the same if the pNet contains no assignment to its state variables.

We define the function $States$ as before, and the function $Assignments : \overline{OT}\ \times State\rightarrow \overline{Assign}$
computing the set of assignments of a global state from a given set of OTs.

The structure implicitly computed by both algorithms is a triple $ Aut= \langle \overline{OT}, \overline{State},\\ \overline{Assign}\rangle$.

Then the assignments set of $OA_1$ is $\{Assignments(L1, S_i)\}^{S_i\in States(L1)}$ and $OA_2$ is $\{Assignments(L2_{sat}, S_j)\}^{S_j\in States(L2_{sat})}$. The result triples are:  
\noindent
\begin{align*}
Aut_1 = \langle L1_{sat}, States(L1_{sat}), \{Assignments(L1, S_i)\}^{S_i\in States(L1)}\rangle\\
Aut_2 = \langle L2_{sat}, States(L2_{sat}), \{Assignments(L2_{sat}, S_j)\}^{S_j\in States(L2_{sat})}\rangle \ 
\end{align*}

Based on this triple structure, we define the function $Refine : Aut \rightarrow Aut$ modeling $makeReachableSubAutomaton(S0,OTs)$. It uses the assignments of the previous step to refine the computed reachable space:
\begin{align*}
Refine(Aut) &=\langle L', States(L'), \{Assignments(L', S_i)\}^{S_i\in States(L')}\rangle \\
&=Aut'
\end{align*}
where $L' = L - L_{unsat}$, and $L_{unsat}$ is the set of unsatisfiable transitions checked in this iteration. We have $L'\subseteq L$, and potentially, their target states and corresponding assignments may be reduced: 
\[States(L')=\{S_0\}\cup \bigcup States(L'),\]
\[Assignments_{L',States_{L'}}=\bigcup\{Assignments(L', State_{i})\}^{State_{i}\in States_{L'}} \]
\[s.t. |States_{L'}|\le|States_{L}|\wedge|Assignments_{L',States_{L'}}|\le|Assignments_{L,States_{L}}|.\]

In one single iteration, satisfiability of a transition $\ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\in L$ is checked through whether 
\[\bigvee_{Assign_i\in Assignments_{L, States_{L}}}(Pred_{l}\wedge Assign_{i})\text{ is true}.\]
More transitions could be unsatisfiable after the reduction of the assignments set, so the function need iterate several times. Given that in each iteration, the number of triples decreases, so $Refine$ is a monotone function. And there exists a sequence \[Aut^n = Refine(Aut^{n-1})\quad s.t. |Aut^n| < |Aut^{n-1}|.\]
The function terminates if $Aut''$ contains no more unsatisfiable transition, then there is a fixed-point $Aut'' = Refine(Aut'')$ and $|Aut''| = |Refine(Aut'')|$.

Let $Aut''_1$ and $Aut''_2$ be two fixed-points that $Refine(Aut''_1) = Aut''_1$, $Refine(Aut''_2) = Aut''_2$. Focusing on the transition sets in $Aut''_1$ and $Aut''_2$, $L1''_{sat}=L1_{reach}$ and $L2''_{sat}=L2_{reach}$, we need to check their equivalence.

@@

Let $L1'_{sat}=L1_{sat}-L1_{unsat}$ and $L2'_{sat}=L2_{sat}-L2_{unsat}$
Then in the next iteration, the unsatisfiable transition $\ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}$ removed from $L1'_{sat}$ might be the same transition (up to renaming) removed from $L2'_{sat}$. In this case,
\[\ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\in L1'_{unsat} \longrightarrow \ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\in L2'_{unsat}.\]
Or it is the transition has not removed in the last iteration because of 
\[Assignments_{L2_{sat}, States_{L2_{sat}}}\subseteq Assignments_{L1, States_{L1}}.\] 
We have already proved that 
\[Assignments_{L1'_{sat}, States_{L1'_{sat}}}\subseteq Assignments_{L2_{sat}, States_{L2_{sat}}},\]
so those transitions will be removed this time. In this case,
\[\ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\in L1'_{unsat} \longrightarrow \ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\notin L2_{sat}.\]

@@

We check the satisfiability of each open transitions
using the SMT solver Z3.
%More precisely, we translate
%the OT predicate into an SMTlib assertion in the context provided by
%the possible state variable assignments in the source state.
In this
section, we describe the translation of the algebra presentation, of
assignments, and of the predicates.

Our implementation submits satisfiability requests to
Z3 using its Java API. Here, for readability, we
show the Z3 code using its SMTlib input language.
Note that in the previous sections, the OTs were displayed in a
simplified, human readable form. The input and output of our tool, and
also the generated SMTlib fragments, are
slightly more difficult to read, in 
particular because of structured names for the fresh variables generated
by the algorithm, allowing for tracability of the result~\cite{Avocs-RR}.

Figure~\ref{schema:smt-lib} shows the translation of  Figure~\ref{schema:unsat-ot}
in the SMTlib syntax. It contains the declaration
of the relevent part of the Lotos action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.

@@

For simplicity, we have only shown here a use-case with boolean conditions, so we only require the basic SAT capability of the Z3
engine. It should be clear that more involved examples will require
reasoning about predicates of various data types. Z3 provides theories for reasonning
about integers, bitstrings, enumerated types, arrays, etc.
We also need to deal with some universal quantification on free
variables occuring in the guards of synchronisation vectors.
For other data structures  in such models, like records, or
recursive structures, the user will have to develop its own theories.
Decidability of these theories will condition on the completeness of the
satisfiability check, and the decidability of model-checking or
bisimulation checking.

@@

\emph{Translation of action algebra presentation.}
In~\cite{AVOCS18}, we  define the translation of an algebra presentation into
SMTlib declarations (\texttt{declare-datatypes} and
\texttt{declare-fun}). We also formalise a condition of well-formedness of pNets to ensure that the generated code is correct
and will not raise runtime errors in the SMT engine. Note that the
\texttt{declare-datatypes} command comprises both the action
constructors from Table~\ref{table:BIPalgebra} and also the constant action
names from our example.
In addition, we will include axiomatisation of any required functions
and predicates of the presentation data-types.

@@

\emph{Translation of open transitions.}
In~\cite{Avocs-RR} we formally define all steps of the translation of each
open transition, including:
%
\begin{itemize}
	\item to collect all variables in the transition, and declare them (using \texttt{declare-const})
	\item to check the well-formedness and correct typing of expressions,
	\item to translate the predicate into a conjunction of assertions,
	\item to translate the state-variable
	assignments into a disjunctive assertion, if the assignments are present in the source state.
\end{itemize}
%
This translation ensures that no runtime error will occur
in the SMT engine.

@@

\section{Use-Cases}
\label{section:use-cases}
In this section we present the results of open automata construction
for several examples, starting with three constructions from
Section~\ref{section:running-example} using the Enable operator.
Then we give detailed results for the Failure Controller that was
presented in \cite{AVOCS18}, and show how this sort of BIP Architectures
can be composed together.
Finaly we give some figures illustrating the benefits from the smart
algorithm on these examples.

@@

For this example, the tool (using Algorithm 1) builds 408 open transitions, whereof 396
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 12 open
transitions, is shown in Figure~\ref{schema:resultOA1}.
%% \noteSB{Eric, in the state 20, $b0$ should always be $false$: $\PortFail(true)$ is not possible in Control, and the safety property (\ref{eq:FM:safety}) does not hold otherwise.}

>>>>

pNets can encode naturally the notion of input actions as found,
\eg in value-passing CCS 
\cite{Milner89} or of usual point-to-point message passing calculi,
but it also allows 
for more general mechanisms, like gate negotiation in LOTOS~\cite{LotosISO89}, or broadcast
communications.

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$
          Add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{AO_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$; \Comment Fixpoint reached
\State \Return $\mathit{OA_n}$;

@@

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
\State $\mathit{L2}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L2)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{AO_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$;
\State \Return $\mathit{OA_n}$;
%\EndIf
\end{algorithmic}  
\end{algorithm}

@@

 Let $pn=$ be a pNet. The set of states 
 managed by the two algorithms both are subsets of the cartesian
 product of the PLTS states, as defined in page \pageref{def-states}. In a
 similar way, the set of state variables attached to each state are
 also equal. Comparing  
 transitions requires a little more care, because within the
 $MakeTransitions$ function, fresh variables are created everytime a
 synchronisation vector is used (this occurs within the $MkPred$ and
 $matching$ functions). However, given a global state, the call of
 $MakeTransitions$ will always return exactly the same set of OTs, up to a
renaming of these fresh variables. When this is not ambiguous, we will
speak of equality (of transitions, and of open automata) without the phrase ``up to a renaming of fresh variables''.

  Let us first define some notations.
 \begin{itemize}
    \item $L0$: the original set of transitions by Algorithm~\ref{alg1}
      (i.e. before line 11), 
    \item $L1_{sat}$:  the original set of transitions by Algorithm~\ref{alg1}, after
      satisfiability check (i.e. after line 11),
    \item $L1_{reach}$: the final set of OTs of Algorithm~\ref{alg1}, as computed
      by line 12 (as $OA_0$). Without assignments the following
      refinement step will do nothing, the fixpoint is already reached.
    \item $L2_{sat}$: the original set of transitions by Algorithm~\ref{alg3}
      (i.e. before line 13), 
    \item $L2_{reach}$: the final set of OTs of Algorithm~\ref{alg3}, as computed
          by line 13. Here also $OA_n = OA_0$.
   \item $States: \overline{OT}\rightarrow\overline{State}$:  the function that
computes the set of target states of a set of OTs.  
  \item $Rename:  L1_{reach} \rightarrow L2_{reach}$:  a function that keeps state names of transitions but renames fresh variables in the transition labels.
\end{itemize}

%% When the pNets contains no assignment, there is no need for the last
%% filtering step, so $L1_{sat}=L1_{reach}$ and $L2_{sat}=L2_{reach}$.

\smallskip
We now prove that $L1_{reach}=L2_{reach}$, which will imply
the equality of the two open automata.

\smallskip
Let $S=\ostate{\overline{s}}=\ostate{\{s_{i}\}^{i\in Leaves(pn)}}$
range over the global states $States(L0)$, 
The only difference between the main loops of the algorithms is that 
Algorithm~\ref{alg3} only retains the satisfiable OTs (lines 6-7), so 
Algorithm~\ref{alg3} computes fewer transitions than
Algorithm~\ref{alg1}. Formally, we have that $L2_{reach}\subseteq  L2_{sat} \subseteq L0$, and as a consequence $States(L2_{reach})\subseteq  States(L0)$, i.e.,\\
\centerline{$\forall \ostate{\overline{u}}\in States(L2_{reach}), \exists\ostate{\overline{t}}\in States(L0), \ostate{\overline{t}}=\ostate{\overline{u}}$}

For a global state $S$, and a set of OTs $L$, define its distance
$|S|_L$ as the length $n$ of (one of the) shortest chain
$S_0\xrightarrow{l_1}S_1\xrightarrow{l_2}...S_{n-1}\xrightarrow{l_{n}}S$
  from the initial state $S_0$, with $l_k \in L$ for each $k\in [1..n]$.

\medskip
We will prove by induction on the distance from the original state,
that $\forall S \in L0. \forall n \in \mathbb{N}. S \in States(L2_{reach}) \land |S|_{l2_{reach}}=n => S
\in States(L1_{reach})$, and vice-versa.

@@

  Suppose our property is true for distance $n$, and
  suppose $|S|_{L2_{reach}}=n+1$, then there exists a chain
  $S_0\xrightarrow{l_1}S_1\xrightarrow{l_2}...S_{n}\xrightarrow{l_{n+1}}S$
  with all $l_k \in L2_{reach}$. In particular, since
  $L2_{reach}\subseteq L2_{sat}$, we have that $l_{n+1} \in L2_{sat}$ and
  its predicate is satisfiable.

  By induction hypothesis, we get $S_{n} \in States(L1_{reach})$, so during the
  call of $makeReachableSubAutomaton$ in Algorithm~\ref{alg1}, line 12, the OT
  $S_n \xrightarrow{l_{n}}S$ will be checked and found SAT, so
  $S \in States(L1_{reach})$.

  Conversely, if $|S|_{L1_{reach}}=n+1$, we get a similar chain in
  $L1_{reach}$, with the transition $S_n \xrightarrow{l_{n}}S$.  Then $ S_n \in
  States(L1_{reach})$ and $l_n$ is satisfiable. By induction, we have
  $S_n \in States(L2_{reach})=States(L2_{sat})$, so $l_n$ will have
  been checked ``on-the-fly'', in line 6 of Algorithm~\ref{alg3}. Then
  the transition $S_n \xrightarrow{l_{n}}S$ will be added to $L2_{sat}$, and thus $S \in States(L2_{reach})$.

Note that $L0$ is finite, so we conclude $L1_{reach}=L2_{reach}$, and $States(L1_{reach})=States(L2_{reach})$.

@@

\paragraph{General case with assignments.}
In the presence of assignments, we need to collect all of them
  before using the assignment information to refine the SAT
  checking. Still, the smart strategy will explore a reduced state
  space, not necessarily the minimal one. In the
  following theorem, both algorithms are considered in their complete
  version, meaning the reachability (with assignments) step is
  iterated up to fixed points.

\begin{theorem}[Correctness with assignments]
  \label{Th:CorrectnessWithAssignments}
In the general case, if a pNet (verifying the finiteness conditions of Theorem 1), contains some assignments to state variables, the smart algorithm computes 
the same open automaton as that of Algorithm 1 up to a renaming of variables in the transition labels.
\end{theorem}

%\TODO{Formalize the proof}
\begin{proof}
  Define the OT sets $L0, L1_{sat}, L2_{sat}, L2_{reach}$ and the function $States$ as in the proof of Theorem~\ref{Th:CorrectnessWithoutAssignments}, and

  \begin{itemize}
    \item $L1_{reach*}$: the final set of OTs of Algorithm~\ref{alg1}, as computed
        by an iterative application of $makeReachableSubAutomaton$ (line
        14) up to a  fixed point.
      \item $L2_{reach*}$: the final set of OTs of Algorithm~\ref{alg3}, as computed
        by an iterative application of $makeReachableSubAutomaton$ (line
        16) up to a fixed point.
        \item $Assignments : \overline{OT}\ \times State\rightarrow \overline{Assign}$: the function of
        computing the set of assignments of a global state from a given set of OTs.
  \end{itemize}

%?? In Theorem 2, we have already proved that the result automata $OA_1$ and $OA_2$ of the two algorithms are the same if the pNet contains no assignment to its state variables.

\smallskip
The structure implicitly computed by both algorithms is a triple $ Aut= \langle \overline{OT}, \overline{State},\\ \overline{Assign}\rangle$.
The assignments set of $OA_1$ is $\{Assignments(L1, S_i)\}^{S_i\in States(L1)}$ and that of $OA_2$ is $\{Assignments(L2_{sat}, S_j)\}^{S_j\in States(L2_{sat})}$. The result triples are:  
\noindent
\begin{align*}
Aut_1 = \langle L1_{sat}, States(L1_{sat}), \{Assignments(L1, S_i)\}^{S_i\in States(L1)}\rangle\ , \\
Aut_2 = \langle L2_{sat}, States(L2_{sat}), \{Assignments(L2_{sat}, S_j)\}^{S_j\in States(L2_{sat})}\rangle \  .
\end{align*}

Based on this triple structure, we define the function $Refine : Aut \rightarrow Aut$ modelling $makeReachableSubAutomaton(S0,OTs)$. It uses the assignments of the previous step to refine the computed reachable space:
\begin{align*}
Refine(Aut) &=\langle L', States(L'), \{Assignments(L', S_i)\}^{S_i\in States(L')}\rangle \\
&=Aut'
\end{align*}
where $L' = L - L_{unsat}$, and $L_{unsat}$ is the set of unsatisfiable transitions checked in this iteration. We have $L'\subseteq L$, and potentially, their target states and corresponding assignments may be reduced: 
\[States(L')=\{S_0\}\cup \bigcup States(L'),\]
\[Assignments_{L',States_{L'}}=\bigcup\{Assignments(L', State_{i})\}^{State_{i}\in States_{L'}} \]
\[s.t. |States_{L'}|\le|States_{L}|\wedge|Assignments_{L',States_{L'}}|\le|Assignments_{L,States_{L}}|.\]

In one single iteration, satisfiability of a transition $\ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\in L$ is checked through whether 
\[\bigvee_{Assign_i\in Assignments_{L, States_{L}}}(Pred_{l}\wedge Assign_{i})\text{ is true}.\]
More transitions could be unsatisfiable after the reduction of the assignments set, so the function needs to be iterated several times. In each iteration, the number of triples decreases, thus there exists a sequence \[Aut^n = Refine(Aut^{n-1})\quad s.t. |Aut^n| < |Aut^{n-1}|.\]
The iteration terminates if $Aut''$ contains no more unsatisfiable transition, then there is a fixed point $Aut'' = Refine(Aut'')$ and $|Aut''| = |Refine(Aut'')|$.

Let $Aut''_1$ and $Aut''_2$ be two fixed points with $Refine(Aut''_1) = Aut''_1$ and $Refine(Aut''_2) = Aut''_2$. Focusing on the transition sets in $Aut''_1$ and $Aut''_2$, $L1''_{sat}=L1_{reach}$ and $L2''_{sat}=L2_{reach}$, we need to check their equivalence.

@@

Let $L1'_{sat}=L1_{sat}-L1_{unsat}$ and $L2'_{sat}=L2_{sat}-L2_{unsat}$
Then in the next iteration, the unsatisfiable transition $\ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}$ removed from $L1'_{sat}$ might be the same transition (up to renaming) removed from $L2'_{sat}$. In this case,
\[\ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\in L1'_{unsat} \longrightarrow \ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\in L2'_{unsat}.\]
Or it is the transition not removed in the last iteration because of the inclusion
\[Assignments_{L2_{sat}, States_{L2_{sat}}}\subseteq Assignments_{L1, States_{L1}}.\] 
We have already proved that 
\[Assignments_{L1'_{sat}, States_{L1'_{sat}}}\subseteq Assignments_{L2_{sat}, States_{L2_{sat}}},\]
so those transitions will be removed this time. In this case,
\[\ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\in L1'_{unsat} \longrightarrow \ostate{\tilde{t}}\xrightarrow{l}\ostate{\tilde{t'}}\notin L2_{sat}.\]

@@

We check the satisfiability of each open transition
using the SMT solver Z3.
%More precisely, we translate
%the OT predicate into an SMTlib assertion in the context provided by
%the possible state variable assignments in the source state.
In this
section, we describe the translation of the algebra presentation, of
assignments, and of the predicates.

Our implementation submits satisfiability requests to
Z3 using its Java API. Here, for readability, we
show the Z3 code using its SMTlib input language.
Note that in the previous sections, the OTs were displayed in a
simplified, human readable form. The input and output of our tool, and
also the generated SMTlib fragments, are
slightly more difficult to read, in 
particular because of structured names for the fresh variables generated
by the algorithms, allowing for tracability of the result~\cite{Avocs-RR}.

Figure~\ref{schema:smt-lib} shows the translation of  Figure~\ref{schema:unsat-ot}
in the SMTlib syntax. It contains the declaration
of the relevent part of the LOTOS action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.

@@

For simplicity, we have only shown here a use-case with boolean conditions, so we only require the basic SAT capability of the Z3
engine. It should be clear that more involved examples will require
reasoning about predicates of various data types. Z3 provides theories for reasonning
about integers, bitstrings, enumerated types, arrays, etc.
We also need to deal with some universal quantification on free
variables occuring in the guards of synchronisation vectors.
For other data structures  in such models, like records, or
recursive structures, the user will have to develop her own theories.
Decidability of these theories will condition on the completeness of the
satisfiability check, and the decidability of model-checking or
bisimulation checking.

@@

\emph{Translation of action algebra presentation.}
In~\cite{AVOCS18}, we  define the translation of an algebra presentation into
SMTlib declarations (\texttt{declare-datatypes} and
\texttt{declare-fun}). We also formalise a condition of well-formedness of pNets to ensure that the generated code is correct
and will not raise runtime errors in the SMT engine. Note that the
\texttt{declare-datatypes} command comprises both the action
constructors from Table~\ref{table:BIPalgebra} and also the constant action
names from our example.
In addition, we will include the axiomatisation of any required functions
and predicates of the presentation data-types.

@@

\emph{Translation of open transitions.}
In~\cite{Avocs-RR} we formally define all steps of the translation of each
open transition, including:
%
\begin{itemize}
	\item to collect all variables in the transition, and declare them (using \texttt{declare-const}),
	\item to check the well-formedness and correct typing of expressions,
	\item to translate the predicate into a conjunction of assertions,
	\item to translate the state-variable
	assignments into a disjunctive assertion, if the assignments are present in the source state.
\end{itemize}
%
This translation ensures that no runtime error will occur
in the SMT engine.

@@

\section{Use-Cases}
\label{section:use-cases}
In this section we present the results of open automata construction
for several examples, starting with three constructions from
Section~\ref{section:running-example} using the Enable operator.
Then we give detailed results for the Failure Controller that was
presented in \cite{AVOCS18}, and show how this sort of BIP Architectures
can be composed together.
Finally we give some figures illustrating the benefits of the smart
algorithm on these examples.

@@

For this example, the tool (using Algorithm~\ref{alg1}) builds 408 open transitions, whereof 396
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 12 open
transitions, is shown in Figure~\ref{schema:resultOA1}.
%% \noteSB{Eric, in the state 20, $b0$ should always be $false$: $\PortFail(true)$ is not possible in Control, and the safety property (\ref{eq:FM:safety}) does not hold otherwise.}
